{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-11-11 ms.tex
Add references to other papers in same issue (requested by editors)

Plant growth and demography are fundamentally size- and
trait-structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}.
%
As an individual plant increases its leaf area, it increases its potential
to generate photosynthate.
%
On the other hand, as individuals grow larger, they must allocate increasing
fractions of their photosynthetic income to activities other
than building new leaves; for example, to maintaining support tissues
\citep{Givnish-1988, Enquist-2007} or to reproduction
\citep{Thomas-2011}.
%
Consequently, rates of growth, mortality and reproduction change with
individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
Ontogenetic
patterns of growth also vary with traits; for example,
leaf and wood construction costs influence growth rates
\citep{Wright-2010}, while seed size and height
at maturation determine start and end points of ontogenetic
trajectories \citep{Westoby-2002}.
%
Strong feedbacks emerge between individuals within a forest via
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals \citep{Shugart-1980, Pacala-1996}.
%
Such feedbacks make it difficult to link observable species traits to
phenomena such as self-thinning, successional transitions, trait
evolution and species coexistence, without
modelling both individual growth and the competitive interactions between
individuals.

Although the importance of size structure for trait evolution, vegetation
dynamics and diversity has long been recognised \citep[e.g.,][]{Harper-1977,
 Shugart-1980, Huston-1987}, current research in this area is
dominated by models and theory that either omit size structure
entirely, or only include coarse approximations.
%
Theoretical investigations exploring questions about niche-based
species coexistence and ecological drift (neutrality) both rely
primarily on unstructured population models \citep[e.g.,][]{MacArthur-1967,
 Tilman-1985, Geritz-1998, Hubbell-2001, Calcagno-2006}.
%
It is common to assume competitive outcomes to be influenced by
size-related traits such as adult size and seed size, but detailed
size structure is rarely considered, at least in plant models
\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
%
Similarly, many of the models used to study global vegetation dynamics across
the last 20 years have not included size-structured demography
\citep[for comparisons of some major models, see][]{Sitch-2008,
Dekauwe-2014}.
%
While the importance of explicitly including size-structured dynamics is
increasingly recognised in both evolutionary ecology
\citep[e.g.,][]{Falster-2015} and vegetation dynamics
\citep[e.g.,][]{Moorcroft-2001, Purves-2008, Smith-2014,
Weng-2015, Sakschewski-2015}, our understanding of how these features
impact the ecology and evolution of vegetation communities remains
relatively limited.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of metapopulations structured by a
disturbance regime.
%
Our own purpose in developing {\plant} has been mainly to investigate how species
differing in traits may be able to coexist with one another
\citep[following ][]{Falster-2011, Falster-2015}; as size-structured models
provide unique opportunities for studying coexistence via differentiation in
successional strategy \citep[see also][]{Huston-1987, Moorcroft-2001}.
At the same time, we expect {\plant} to be useful for modelling other demographic and evolutionary phenomena.

@@

Within patches of competing plants, competition for light generates
strong nonlinear feedbacks on growth, survival and reproduction. In the
\texttt{FF16} physiological model, we consider only the effects of
shading on rates of biomass production. Competition for other
resources such as nitrogen or water could be considered by extending the
physiological model. We assume that patches are vertically -- but not horizontally --
structured; in other words, we account for size differences, but not for the spatial
layout within patches. Similar assumptions are made in other \textsc{TSPMs}
\citep{Shugart-1980, Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}.
While one would ideally also consider spatial interactions within patches,
such models are very computationally demanding \citep{Shugart-1980,
Pacala-1996}. Simplifying spatial interactions within patches can be motivated from the observation
that competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

@@

Our approach for numerically solving deterministic size-structured population dynamics is
based on the characteristic method \citep{Angulo-2004}. This method describes the development of
the size distribution\(N\) by approximating it along a collection of individual growth trajectories spanning the size
spectrum. Following a disturbance, a series of such trajectories are introduced
into each patch. These trajectories then change according to the growth
of the corresponding individuals, conditioned on their survival
(Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
to alter the density of individuals along these trajectories
(Fig. \ref{fig:patch}b). The characteristic method is similar to the Escalator Boxcar Train (\textsc{EBT})method
\citep{Deroos-1988, Deroos-1997} used by \citet{Falster-2011} and
\citet{Falster-2015}, but not identical. An advantage of the characteristic method is that it allows direct approximation of the size distribution. For the same reason, however, the characteristic method may be less suitable for resolving size distributions in models where those distributions naturally develop sharp peaks (which happens, e.g., when plants completely cease to grow at a given height).

@@

All vegetation in {\plant} is subject to a disturbance regime, implying that patches
are cleared at a given rate. Such clearance can be interpreted, for example
as resulting from a fire, cyclone, clear-cutting or
landslide. The vegetation thus comprises a series of patches
differing in the time that has passed since their last disturbance, linked via seed dispersal
(Fig. \ref{fig:schematic}b). Such a system is often referred to as a
structured metapopulation \citep{Gyllenberg-2001}.

@@

Increased interest in capturing the dynamics of size-structured populations
has prompted development of several tools, each with its own distinct
strengths. The \textsc{EBT} software (see De Roos 2016 in this special feature) provides
a flexible framework that has been used to study a range of organisms.
Compared to the \textsc{EBT}, {\plant} is customised for plants, and
to that end offers refinements needed to handle modelling of the strongly
size-asymmetric feedbacks that occur in plant communities, as well as the
extension to a structured metapopulation. Other \textsc{TSPMs} also exist for modelling of vegetation, like \textsc{ED}
\citep[ver 1 and 2;][]{Moorcroft-2001, Medvigy-2009}, \textsc {LPJ-GUESS}
\citep{Smith-2014} and \textsc{LPJmL-FIT} \citep{Sakschewski-2015}. These
models were designed for simulating bio-geochemical cycles and vegetation
structure across broad geographic areas, and their architectures are
optimised for such use. By contrast, {\plant} was designed so that the
physiological rules driving the dynamics of the system could be easily
manipulated and emergent demographic properties (including those of individual plants, patches, metapopulations of patches and fitness) studied in
detail. Finally, like {\plant}, the \textsc{LM3-PPA} \cite{Weng-2015} model also aims
to scale from individual-level processes to emergent properties of
vegetation, including estimating invasion fitness. Yet the approaches taken
for solving the dynamics are very different. The \textsc{LM3-PPA} relies on
complex analytical solutions to the size-structured interactions, whose
derivation required simplifying elements of the physiology and ecology in that
model. By contrast, in {\plant} we have retained the full range of
physiological and ecological feedbacks, and instead relied on using robust
numerical methods to solve the system dynamics (see Appendix
\ref{sec:demography} for details).

>>>>

Plant growth and demography are fundamentally size- and
trait-structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002, Griffith-2016, Rees-2016}.
%
As an individual plant increases its leaf area, it increases its potential
to generate photosynthate.
%
On the other hand, as individuals grow larger, they must allocate increasing
fractions of their photosynthetic income to activities other
than building new leaves; for example, to maintaining support tissues
\citep{Givnish-1988, Enquist-2007} or to reproduction
\citep{Thomas-2011}.
%
Consequently, rates of growth, mortality and reproduction change with
individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
Ontogenetic
patterns of growth also vary with traits; for example,
leaf and wood construction costs influence growth rates
\citep{Falster-2011, Visser-2016}, while seed size and height
at maturation determine start and end points of ontogenetic
trajectories \citep{Westoby-2002}.
%
Strong feedbacks emerge between individuals within a forest via
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals \citep{Shugart-1980, Pacala-1996}.
%
Such feedbacks make it difficult to link observable species traits to
phenomena such as self-thinning, successional transitions, trait
evolution and species coexistence, without
modelling both individual growth and the competitive interactions between
individuals.

Although the importance of size structure for trait evolution, vegetation
dynamics and diversity has long been recognised \citep[e.g.,][]{Harper-1977,
 Shugart-1980, Huston-1987}, current research in this area is
dominated by models and theory that either omit size structure
entirely, or only include coarse approximations.
%
Theoretical investigations exploring questions about niche-based
species coexistence and ecological drift (neutrality) both rely
primarily on unstructured population models \citep[e.g.,][]{MacArthur-1967,
 Tilman-1985, Geritz-1998, Hubbell-2001, Calcagno-2006}.
%
It is common to assume competitive outcomes to be influenced by
size-related traits such as adult size and seed size, but detailed
size structure is rarely considered, at least in plant models
\cite[for animal examples, see][]{Deroos-1997}.
%
Similarly, many of the models used to study global vegetation dynamics across
the last 20 years have not included size-structured demography
\citep[for comparisons of some major models, see][]{Sitch-2008,
Dekauwe-2014}.
%
While the importance of explicitly including size-structured dynamics is
increasingly recognised in both evolutionary ecology
\citep[e.g.,][]{Falster-2015, Rees-2016} and vegetation dynamics
\citep[e.g.,][]{Moorcroft-2001, Purves-2008, Smith-2014,
Weng-2015, Sakschewski-2015}, our understanding of how these features
impact the ecology and evolution of vegetation communities remains
relatively limited.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of metapopulations structured by a
disturbance regime.
%
Our own purpose in developing {\plant} has been mainly to investigate how species
differing in traits may be able to coexist with one another
\citep[following ][]{Falster-2011, Falster-2015}; as size-structured models
provide unique opportunities for studying coexistence via differentiation in
successional strategy \citep[see also][]{Huston-1987, Moorcroft-2001, Uriate-2016}.
At the same time, we expect {\plant} to be useful for modelling other demographic and evolutionary phenomena.

@@

Within patches of competing plants, competition for light generates
strong nonlinear feedbacks on growth, survival and reproduction. In the
\texttt{FF16} physiological model, we consider only the effects of
shading on rates of biomass production. Competition for other
resources such as nitrogen or water could be considered by extending the
physiological model. We assume that patches are vertically -- but not horizontally --
structured; in other words, we account for size differences, but not for the spatial
layout within patches. Similar assumptions are made in other \textsc{TSPMs}
\citep{Shugart-1980, Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}, and also
in related approaches, such as Integral Projection Models \citep[e.g.,][]{Rees-2016}.
While one would ideally also consider spatial interactions within patches,
such models are very computationally demanding \citep{Shugart-1980,
Pacala-1996}. Simplifying spatial interactions within patches can be motivated from the observation
that competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

@@

Our approach for numerically solving deterministic size-structured population dynamics is
based on the characteristic method \citep{Angulo-2004}. This method describes the development of
the size distribution\(N\) by approximating it along a collection of individual growth trajectories spanning the size
spectrum. Following a disturbance, a series of such trajectories are introduced
into each patch. These trajectories then change according to the growth
of the corresponding individuals, conditioned on their survival
(Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
to alter the density of individuals along these trajectories
(Fig. \ref{fig:patch}b). The characteristic method is similar to the Escalator Boxcar Train (\textsc{EBT})method
\citep{Deroos-1997} used by \citet{Falster-2011} and
\citet{Falster-2015}, but not identical. An advantage of the characteristic method is that it allows direct approximation of the size distribution. For the same reason, however, the characteristic method may be less suitable for resolving size distributions in models where those distributions naturally develop sharp peaks (which happens, e.g., when plants completely cease to grow at a given height).

@@

All vegetation in {\plant} is subject to a disturbance regime, implying that patches
are cleared at a given rate \citep[e.g.,][this issue]{Treurnicht-2016}.
Such clearance can be interpreted, for example
as resulting from a fire, cyclone, clear-cutting or
landslide. The vegetation thus comprises a series of patches
differing in the time that has passed since their last disturbance, linked via seed dispersal
(Fig. \ref{fig:schematic}b). Such a system is often referred to as a
structured metapopulation \citep{Gyllenberg-2001}.

@@

Increased interest in capturing the dynamics of size-structured populations
\citep[e.g.][]{Griffith-2016}
has prompted development of several tools, each with its own distinct
strengths. The \textsc{EBT} software \citep{Deroos-1997}, {\plant} is customised for plants, and
to that end offers refinements needed to handle modelling of the strongly
size-asymmetric feedbacks that occur in plant communities, as well as the
extension to a structured metapopulation. Other \textsc{TSPMs} also exist for modelling of vegetation, like \textsc{ED}
\citep[ver 1 and 2;][]{Moorcroft-2001, Medvigy-2009}, \textsc {LPJ-GUESS}
\citep{Smith-2014} and \textsc{LPJmL-FIT} \citep{Sakschewski-2015}. These
models were designed for simulating bio-geochemical cycles and vegetation
structure across broad geographic areas, and their architectures are
optimised for such use. By contrast, {\plant} was designed so that the
physiological rules driving the dynamics of the system could be easily
manipulated and emergent demographic properties (including those of individual plants, patches, metapopulations of patches and fitness) studied in
detail. Finally, like {\plant}, the \textsc{LM3-PPA} \cite{Weng-2015} model also aims
to scale from individual-level processes to emergent properties of
vegetation, including estimating invasion fitness. Yet the approaches taken
for solving the dynamics are very different. The \textsc{LM3-PPA} relies on
complex analytical solutions to the size-structured interactions, whose
derivation required simplifying elements of the physiology and ecology in that
model. By contrast, in {\plant} we have retained the full range of
physiological and ecological feedbacks, and instead relied on using robust
numerical methods to solve the system dynamics (see Appendix
\ref{sec:demography} for details).
